Investigating the Molecular Processes Behind the Cell-Specific Toxicity Response to Titanium Dioxide Nanobelts

### 2021-08-26 manuscript.tex
Copied in the proof version from MDPI

%=================================================================
\documentclass[ijms,article,submit,moreauthors,pdftex]{Definitions/mdpi}
%\usepackage[table,xcdraw]{xcolor}

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2021}
\copyrightyear{2020}
%\externaleditor{Academic Editor: Firstname Lastname} % For journal Automation, please change Academic Editor to "Communicated by"
\datereceived{} 
\dateaccepted{} 
\datepublished{} 
\hreflink{https://doi.org/} % If needed use \linebreak
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Investigating the molecular processes behind the cell-specific toxicity response to titanium dioxide nanobelts}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Investigating the molecular processes behind the cell-specific toxicity response to titanium dioxide nanobelts}

@@

% Authors, for the paper (add full first names)
\Author{Laurent A Winckers $^{1}$\orcidA{}, Chris T Evelo $^{1,2}$\orcidB{}, Egon L Willighagen$^{1}$\orcidC{} and Martina Kutmon $^{1,2}$*\orcidD{}}

@@

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Winckers, L.A,; Evelo, C.T; Willighagen, E.L.; Kutmon, M.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Bioinformatics - BiGCaT, NUTRIM School of Nutrition and Translational Research in Metabolism, Maastricht University\\
$^{2}$ \quad Maastricht Centre for Systems Biology (MaCSBio), Maastricht University}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Some engineered nanomaterials incite toxicological effects, but the underlying molecular processes are understudied. The varied physicochemical properties cause different initial molecular interactions, complicating toxicological predictions. Gene expression data allows us to study the responses of genes and biological processes. Overrepresentation analysis identifies enriched biological processes using the experimental data but prompts broad results instead of detailed toxicological processes. We demonstrate a targeted filtering approach to compare public gene expression data for low and high exposure on three cell lines to titanium dioxide nanobelts. Our workflow finds cell and concentration-specific changes in affected pathways linked to four Gene Ontology terms (apoptosis, inflammation, DNA damage and oxidative stress) to select pathways with a clear toxicity focus. 
We saw more differentially expressed genes at higher exposure, but our analysis identifies clear differences between the cell lines in affected processes. Colorectal adenocarcinoma cells showed resilience to both concentrations. Small airway epithelial cells displayed a cytotoxic response to the high concentration, but not as strongly as monocytic like cells. The pathway-gene networks highlighted the gene overlap between altered toxicity-related pathways. The automated workflow is flexible and can focus on other biological processes by selecting other GO terms.}

% Keywords
\keyword{nanomaterials; titanium dioxide; nanobelts; overrepresentation analysis; gene ontology; THP1; SAE; Caco2}

@@

Engineered nanomaterials have become important in our daily life as they are utilized in the fields of food, packaging, cosmetics, drug and vaccine delivery, and many others~\cite{Ray2009}. For example, silver and carbon nanotubes are used in a variety of cleansers because of their antimicrobial properties, and silicon dioxide is used as a food additive as it decreases viscosity and regulates acidity~\cite{Shin2015}. Nevertheless, some nanomaterials, such as asbestos fibres and silica dust, show how small particles can cause adverse outcomes to those exposed to them~\cite{Visona2018,Markowitz2015,Chen2012,Arnoldussen2019}.

The detailed biological processes related to the toxicity of many engineered nanomaterials are not all yet fully understood~\cite{Podila2012,Bahadar2016}. Due to the varied physicochemical properties of nanomaterials and often even of the particles within the nanomaterial itself, it is difficult to predict the biological effects leading to toxicity. Biological response and toxicity depend on the size of the nanoparticle itself, size of the agglomerate, surface impurities, and degradability~\cite{DeStefano2012}. Furthermore, other things that have an effect on the biological response and toxicity include the way of exposure, the entry route into the human body, and the distribution in the body have an effect on the biological response and toxicity~\cite{Shin2015}. The shape of a nanoparticle also has an influence on the nanoparticle’s effects, where for example nanobelts, a nanostructure in the form of a belt, shown to be more pro-inflammatory compared to spheres~\cite{Silva2013}. Studies have found that there are various biological effects and toxicity of engineered nanomaterials under different circumstances and with varying engineered nanomaterials~\cite{Ray2009,Atha2017,Donaldson2003,Sung2008,EbabeElle2013}. These changes in biological effects are ascribed to the differences in chemical and physical properties of nanomaterials. However, studying the property – biological effect relation is complicated because of the difficulty to create identical nanomaterials from different batches and/or sources due to the varying physicochemical properties of nanomaterials~\cite{Mulhopt2018}.Nevertheless, manually curation of transcriptomics datasets regarding nanomaterials are done to enhance their degree of FAIRness to aid nanomaterial safety assessment~\cite{AliisaSaarimaki}.

Furthermore, a relatively well-studied and widely used nanomaterial is titanium dioxide (TiO\textsubscript{2}). Due to its general properties, such as the photocatalytic activity and white color, it is used in various applications such as (photo)catalysis, antibacterial agents, and consumer products. Whereas TiO\textsubscript{2} particles are shown not to be able to penetrate through the skin, entry into the body can happen via inhalation or ingestion where it then has to pass the gastrointestinal tract~\cite{Shakeel2015,Gupta2018}.

Regarding the adverse outcomes, the toxic effects of TiO\textsubscript{2} are known to happen in order, like most toxicity related biological processes. TiO\textsubscript{2} is known to induce reactive oxygen species production which involves lipid peroxidation and eventually will lead to cell damage and DNA damage~\cite{Hou2019,LAzou2008}. Furthermore, The increase in oxidative stress can contribute to the promotion of apoptosis of the affected cells~\cite{Circu2010,Redza-Dutordoir2016}. Consequently, TiO\textsubscript{2} nanoparticles have been shown to induce inflammation~\cite{Yan2020,Baisch2014} among other things via the aforementioned oxidative stress in mammalian cells~\cite{Hou2019,Meena2015}. It has been shown that TiO\textsubscript{2} nanoparticle exposure can lead to impaired immune homeostasis including increases in TNF-$\alpha$, IFN-$\gamma$, IL-2, IL-4, IL-6, IL-8, and IL-10 secretion~\cite{RothenRutishauser2007,Bettini2017,MedinaReyes2015}.

To study the detailed mechanisms of the cellular response, bioinformatics and systems biology approaches, including pathway and network analysis, were used to assess the effects on toxicity-related processes upon exposure to TiO\textsubscript{2}-nanobelts, \textit{i.e.} reactive oxygen species formation and oxidative stress, apoptotic cell death, inflammation, and DNA damage~\cite{Zhang2018,Liu2019}. Pathway enrichment analysis helps to put high-throughput omics data such as transcriptomics into a biological context~\cite{Khatri2012} and the visual diagrams from pathway databases such as WikiPathways~\cite{Martens2020} and Reactome~\cite{Fabregat2017} provide a way to visualize the effects on cellular processes. However, navigating all the affected pathways and the roles of the genes in these pathways can be nontrivial: genes can participate in multiple pathways, and pathways tend to overlap with each other. Moreover, we want to be able to focus on detailed, molecular pathways related to a specific biological process, such as apoptosis, inflammation, DNA damage, and oxidative stress. Overrepresentation, with respect to these processes alone, makes it possible to focus on a subset of genes, but it does not have the link to the pathways. Instead, we selected pathways based on their gene overlap with specific GO-terms covering toxicity-related biological processes and subsequently used the overrepresentation of differentially expressed genes to select relevant pathways out of this subset. We further studied these results in detail using network analysis approaches to investigate pathway overlap and crosstalk.

In this study we re-examined a publicly available dataset generated by Tilton~\textit{et al.}~\cite{Tilton2013} (accession number: GSE42069)~\cite{Edgar2002} to study the detailed molecular mechanisms of TiO\textsubscript{2}-nanobelts toxicity. We analyzed gene expression data from three different cell lines exposed to either one of two concentrations of TiO\textsubscript{2}-nanobelts for 24 hours. The cell lines used were human primary small airway epithelial cells (SAE), human monocytic cells (THP1), and human epithelial colorectal adenocarcinoma cells (Caco2). These cell lines represent three common areas of exposure in the human body \textit{i.e.} skin epithelium, colon epithelium, and the innate immune system that typically will respond to particles that entered the body. We aim to provide an overview of the dose-dependent and cell type-specific response by focusing on toxicity-related processes. Apart from a significant increase in the activity of toxicity-related processes after exposure with a higher dosage, differences in intensity, and affected processes might occur between the cell lines. Using this work as an example, we will discuss the toxicological effects of TiO\textsubscript{2} nanobelts on the three cell lines and the importance and benefits of analysis workflows in the nanomaterial research field.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion}

\subsection*{Differential gene expression}
The differential gene expression analysis for the three cell lines after exposure to two different TiO\textsubscript{2}-nanobelt concentrations show that a high concentration of TiO\textsubscript{2}-nanobelts (Figure~\ref{fig:fig1}, right column) causes stronger gene expression changes in all three cell lines than at a low concentration (Figure~\ref{fig:fig1}, left column). While the Caco2 and SAE cells also show an increased response to the high TiO\textsubscript{2}-nanobelts concentration, the THP1 cells respond much stronger to the high concentration in terms of differentially expressed genes passing our criteria. Dose-dependent increases in response have been shown for many nanomaterials~\cite{Iavicoli2018,Ding2005}.

\begin{figure}[ht!]
\includegraphics[width=0.9\linewidth]{fig1.png}
  \caption{Gene expression volcano plots for different cell lines and TiO\textsubscript{2}-nanobelts concentrations.
   On the x-axis log\textsubscript{2}(fold change) is depicted whereas on the y-axis the -log\textsubscript{10}(p-value) is depicted. The dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ -0.26, p-value $<$ 0.05). Brown color depicts that a gene meets both cut-off criteria, a blue color relates to meeting only the p-value cut-off, an orange color relates to meeting only the log2 fold change cut-off and grey color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}

\subsection*{Pathway analysis}
Using the differentially expressed genes from the different cell lines, overrepresentation analysis was performed to identify altered pathways after TiO\textsubscript{2}-nanobelt exposure. The number of significantly overrepresented pathways (p-value $<$ 0.05) are shown in Table 1 in the column “Significant”. Concordant to the increase in differentially expressed genes matching our criteria shown in Figure 1, the number of resulting pathways increases with a surge in the concentration of TiO\textsubscript{2}-nanobelts. However, there was a smaller increase in differentially expressed genes, we found a similar increase in resulting pathways for the SAE cell line. While there was also an increase in the number of genes found for the Caco2 cell lines it did not directly translate to an increase in overrepresented pathways. Although over-representation analysis provides a great overview of all processes that are affected, it takes time to go manually over the many overrepresented pathways. Therefore, an automated method to select desired pathways, \textit{i.e.} toxicity-related pathways, gives a new approach to interpret the data.

\subsection*{Toxicity-related pathways}
To gain more insight into the toxicity-related processes, the overrepresented pathways were further categorized into pathways related to apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is done manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, \textit{i.e.} “apoptotic process”~(GO:0006915, 1269 genes), “inflammatory response”~(GO:0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO:0006974, 762 genes), and “response to oxidative stress”~(GO:0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap.  Additional file A1 shows the gene overlap between the four GO gene sets in a Venn-diagram (see Additional file A1). Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The overlap was calculated by dividing the number of toxicity-related genes present in the pathway by total number of genes present in the pathway. As a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to GO-term of interest via their gene overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The cut-offs we used were 50\%, 60\%, 70\% and 80\%. For the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The results are shown in Table~A2 (see Additional file A2). The overrepresentation analysis (p-value < $<$ 0.05) showed the most pathways to be linked to the four GO-terms, in this case it showed 401, 250, 203 and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage and oxidative stress GO-terms respectively. The 80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs the at least 50\% gene overlap cut-off was deemed best for our approach.

After this filtering step we found, out of the 1,076 pathways in the human pathway collection, 66 pathways are related to the apoptotic process, 15 are linked to inflammatory response, and 28 are linked to DNA damage and 2 are linked to oxidative stress.
It is noticeable that there are only two oxidative stress pathways, \textit{i.e.} Detoxification of Reactive Oxygen Species~(wikipathways:WP2824~\cite{WP2824}) and Oxidative Stress~(wikipathways:WP408~\cite{WP408}), that have at least 50$\%$ of their genes overlapping with the genes annotated to the GO-term "response to oxidative stress". Based on this finding we can argue that most pathways in the pathway database we have used describe different processes which result in oxidative stress or where oxidative stress is part of/has an influence on the process itself. Whereas these do not describe oxidative stress as a process.  Nevertheless, these pathways are of interest for our analysis.

\subsection*{Study effect on toxicity-related pathways}
The number of altered toxicity-related pathways is shown in Table 1. The overrepresentation analysis was performed using the differentially expressed genes (log2 fold change lower than -0.26 or greater than 0.26 and p-value lower than 0.05).

The Caco2 cell line shows for both concentration 10 overrepresented pathways. While the SAE cell line shows 3 for the low concentration, which could be an indication of a very little toxic response to the TiO\textsubscript{2}-nanobelts exposure, and 15 for the high concentration. The THP1 cell line shows a clear toxic response activation for the high concentration with 39 overrepresented pathways, and 10 for the low concentration.

\begin{table}[]
\caption{Table depicting the number of significantly overrepresented pathways, altered toxicity pathways and the number of altered toxicity pathways for each GO-term.}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcc|cccc}
 & \textbf{Significant} & \textbf{Toxicity} & \textbf{Apoptosis} & \textbf{Inflammation} & \textbf{\begin{tabular}[c]{@{}c@{}}DNA \\ damage\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Oxidative \\ stress\end{tabular}} \\ \hline
\textbf{Pathways} & 1076 & 101 & 66 & 15 & 28 & 2 \\ \hline
Caco2 low & 56 & 10 & 5 & 0 & 5 & 0 \\
Caco2 high & 58 & 10 & 10 & 1 & 0 & 0 \\ \hline
SAE low & 28 & 3 & 1 & 0 & 2 & 0 \\
SAE high & 50 & 15 & 6 & 1 & 8 & 0 \\ \hline
THP1 low & 38 & 10 & 9 & 4 & 0 & 0 \\
THP1 high & 201 & 39 & 31 & 9 & 3 & 1 
\end{tabular}%
}
\end{table}

Importantly, while molecular pathways describe a process on a detailed level, their boundaries are often not clearly defined. Pathways are not independent, and they interact with each other through shared genes or sub-pathways. Figure~\ref{fig:fig2} shows the gene overlap between the altered toxicity-related pathways in pathway-gene networks and highlights the differences in response between the cell lines. Pathways that cluster together indicate that these pathways depict similar biological processes with a high gene overlap. In the following sections, the biological pathways affected in the different cell lines will be discussed in detail.

\begin{figure}[ht!]
\includegraphics[width=0.9\linewidth]{fig2.pdf}
  \caption{Pathway-gene networks of altered toxicity-related pathways.
   Color of the nodes depict to which GO-term the pathway is affiliated. Orange depicts apoptosis, blue depicts inflammation, pink depicts DNA damage and bordeaux depicts oxidative stress. Gray nodes represent genes. Number (n) represents the number of significantly overrepresented pathways that are depicted in the networks.}
\label{fig:fig2}
\end{figure}

>>>>

%=================================================================
\documentclass[ijms,article,accept,moreauthors,pdftex]{Definitions/mdpi}
%\usepackage[table,xcdraw]{xcolor}

\usepackage{xcolor}
\usepackage{soul}

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2021}
\copyrightyear{2021}
\externaleditor{Academic Editor: \hl{Firstname Lastname} %MDPI: please add this if available.
} % For journal Automation, please change Academic Editor to "Communicated by"
\datereceived{} 
\dateaccepted{} 
\datepublished{} 
\hreflink{https://doi.org/} % If needed use \linebreak
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Investigating the Molecular Processes Behind the Cell-Specific Toxicity Response to Titanium Dioxide Nanobelts}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Investigating the Molecular Processes Behind the Cell-Specific Toxicity Response to Titanium Dioxide Nanobelts}

@@

% Authors, for the paper (add full first names)
\Author{\hl{Laurent A. Winckers} %MDPI: Please carefully check the accuracy of names and affiliations.
 $^{1}$\orcidA{}, Chris \hl{T.} %MDPI: we added dot for middle name abbreviation, please confirm.
 Evelo $^{1,2}$\orcidB{}, Egon \hl{L.} Willighagen $^{1}$\orcidC{} and Martina Kutmon $^{1,2,}$*\orcidD{}}

@@

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Winckers, L.A.; Evelo, C.T.; Willighagen, E.L.; Kutmon, M.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Department of Bioinformatics\hl{---}%MDPI: changed from hyphen into em dash.
BiGCaT, NUTRIM School of Nutrition and Translational Research in Metabolism, Maastricht \hl{University}; %MDPI: PLEASE add city, state abbreviation postcode, country.
 \hl{laurent.winckers@maastrichtuniversity.nl (L.A.W.); chris.evelo@maastrichtuniversity.nl (C.T.E.);  egon.willighagen@maastrichtuniversity.nl (E.L.W.)} %MDPI: Newly added information, please confirm.
\\
$^{2}$ \quad Maastricht Centre for Systems Biology (MaCSBio), Maastricht \hl{University}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Some engineered nanomaterials incite toxicological effects, but the underlying molecular processes are understudied. The varied physicochemical properties cause different initial molecular interactions, complicating toxicological predictions. Gene expression data allow us to study the responses of genes and biological processes. Overrepresentation analysis identifies enriched biological processes using the experimental data but prompts broad results instead of detailed toxicological processes. We demonstrate a targeted filtering approach to compare public gene expression data for low and high exposure on three cell lines to titanium dioxide nanobelts. Our workflow finds cell and concentration-specific changes in affected pathways linked to four Gene Ontology terms (apoptosis, inflammation, DNA damage, and oxidative stress) to select pathways with a clear toxicity focus. 
We saw more differentially expressed genes at higher exposure, but our analysis identifies clear differences between the cell lines in affected processes. Colorectal adenocarcinoma cells showed resilience to both concentrations. Small airway epithelial cells displayed a cytotoxic response to the high concentration, but not as strongly as monocytic-like cells. The pathway-gene networks highlighted the gene overlap between altered toxicity-related pathways. The automated workflow is flexible and can focus on other biological processes by selecting other GO terms.}

% Keywords
\keyword{nanomaterials; titanium dioxide; nanobelts; overrepresentation analysis; Gene Ontology; THP1; SAE; Caco-2/HT29-MTX}

@@

Engineered nanomaterials have become important in our daily life as they are utilized in the fields of food, packaging, cosmetics, drug and vaccine delivery, and~many others~\cite{Ray2009}. For~example, silver and carbon nanotubes are used in a variety of cleansers because of their antimicrobial properties, and~silicon dioxide is used as a food additive as it decreases viscosity and regulates acidity~\cite{Shin2015}. Nevertheless, some nanomaterials, such as asbestos fibers and silica dust, show how small particles can cause adverse outcomes for those exposed to them~\cite{Visona2018,Markowitz2015,Chen2012,Arnoldussen2019}.

The detailed biological processes related to the toxicity of many engineered nanomaterials are not yet all fully understood~\cite{Podila2012,Bahadar2016}. Due to the varied physicochemical properties of nanomaterials and often even of the particles within the nanomaterial itself, it is difficult to predict the biological effects leading to toxicity. Biological response and toxicity depend on the size of the nanoparticle, size of the agglomerate, surface impurities, and~degradability~\cite{DeStefano2012}. Furthermore, other factors that have an effect on biological response and toxicity include the method of exposure, the~entry route into the human body, and~the distribution in the body ~\cite{Shin2015}. %check meaning has been retained - duplicated text removed as well as minor edits to this sentence
The~shape of a nanoparticle also has an influence on the nanoparticle’s effects, where for example nanobelts, nanostructures in the form of belts, have been shown to be more pro-inflammatory compared to spheres~\cite{Silva2013}. Studies have found that there are various biological effects and toxicities of engineered nanomaterials under different circumstances and with varying engineered nanomaterials~\cite{Ray2009,Atha2017,Donaldson2003,Sung2008,EbabeElle2013}. These changes in biological effects are ascribed to the differences in chemical and physical properties of nanomaterials. However, when studying the properties, biological effect relations are complicated because of the difficulty in creating identical nanomaterials from different batches and/or sources due to the varying physicochemical properties of nanomaterials~\cite{Mulhopt2018}. %check meaning has been retained
Nevertheless, manual curation of transcriptomics datasets regarding nanomaterials is performed to enhance their degree of 
fairness  %check meaning has been retained
to aid nanomaterial safety assessment~\cite{AliisaSaarimaki}.

Furthermore, a~relatively well-studied and widely used nanomaterial is titanium dioxide (TiO\textsubscript{2}). Due to its general properties, such as its photocatalytic activity and white color, it is used in various applications such as (photo)catalysis, antibacterial agents, and~consumer products. Whereas TiO\textsubscript{2} particles have been shown not to be able to penetrate through the skin, entry into the body can occur via inhalation or ingestion where it then has to pass through the gastrointestinal tract~\cite{Shakeel2015,Gupta2018}.

Regarding the adverse outcomes, the~toxic effects of TiO\textsubscript{2} are known to occur in order, like most toxicity related biological processes. TiO\textsubscript{2} is known to induce reactive oxygen species production, which involves lipid peroxidation and will eventually lead to cell damage and DNA damage~\cite{Hou2019,LAzou2008}. Furthermore, the increase in oxidative stress can contribute to the promotion of apoptosis of the affected cells~\cite{Circu2010,Redza-Dutordoir2016}. Consequently, TiO\textsubscript{2} nanoparticles have been shown to induce inflammation~\cite{Yan2020,Baisch2014} among other things via the aforementioned oxidative stress in mammalian cells~\cite{Hou2019,Meena2015}. It has been shown that TiO\textsubscript{2} nanoparticle exposure can lead to impaired immune homeostasis including increases in TNF-$\alpha$, IFN-$\gamma$, IL-2, IL-4, IL-6, IL-8, and~IL-10 secretion~\cite{RothenRutishauser2007,Bettini2017,MedinaReyes2015}.

To study the detailed mechanisms of the cellular response, bioinformatics and systems biology approaches, including pathway and network analysis, have been used to assess the effects on toxicity-related processes upon exposure to TiO\textsubscript{2}-nanobelts, i.e.,~reactive oxygen species formation and oxidative stress, apoptotic cell death, inflammation, and~DNA damage~\cite{Zhang2018,Liu2019}. Pathway enrichment analysis helps to put high-throughput omics data such as transcriptomics into a biological context~\cite{Khatri2012} and the visual diagrams from pathway databases such as WikiPathways~\cite{Martens2020} and Reactome~\cite{Fabregat2017} provide a way to visualize the effects on cellular processes. However, navigating all the affected pathways and the roles of the genes in these pathways can be nontrivial: genes can participate in multiple pathways, and~pathways tend to overlap with each other. Moreover, we want to be able to focus on detailed, molecular pathways related to a specific biological process, such as apoptosis, inflammation, DNA damage, and~oxidative stress. Overrepresentation, with~respect to these processes alone, makes it possible to focus on a subset of genes, but~it does not have the link to the pathways. Instead, we selected pathways based on their gene overlap with specific GO-terms covering toxicity-related biological processes and subsequently used the overrepresentation of differentially expressed genes to select relevant pathways out of this subset. We further studied these results in detail using network analysis approaches to investigate pathway overlap and~crosstalk.

In this study we re-examined a publicly available dataset generated by Tilton~{et al.}~\cite{Tilton2013} (accession number: GSE42069)~\cite{Edgar2002} to study the detailed molecular mechanisms of TiO\textsubscript{2}-nanobelt toxicity. We analyzed gene expression data from three different {cellular models} exposed to either one of two concentrations of TiO\textsubscript{2}-nanobelts for 24 h. The~cell lines used were human primary small airway epithelial (SAE) cells, human monocytic cells (THP1), and~human epithelial colorectal adenocarcinoma cells {co-cultured with HT29-MTX goblet cells (Caco-2/HT29-MTX)}. These {cellular models} represent three common areas of exposure in the human body, i.e.,~{small airway} epithelium, colon epithelium, and~the innate immune system that will typically respond to particles that have entered the body. We aim to provide an overview of the dose-dependent and cell-type-specific response by focusing on toxicity-related processes. Apart from a significant increase in the activity of toxicity-related processes after exposure with a higher dosage, differences in intensity and~affected processes might occur between the cell lines. Using this work as an example, we will discuss the toxicological effects of TiO\textsubscript{2}-nanobelts on the three {cellular models} and the importance and benefits of analysis workflows in the nanomaterial research~field.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and~Discussion}
\unskip

\subsection{Differential Gene Expression}
The differential gene expression analysis for the three cell lines after exposure to two different TiO\textsubscript{2}-nanobelt concentrations shows that a high concentration of TiO\textsubscript{2}-nanobelts (Figure~\ref{fig:fig1}, right column) causes stronger gene expression changes in all three {cellular models} than at a low concentration (Figure~\ref{fig:fig1}, left column). While the {Caco-2/HT29-MTX} and SAE cells also show an increased response to the high TiO\textsubscript{2}-nanobelt concentration, the~THP1 cells respond much more strongly to the high concentration in terms of differentially expressed genes passing our criteria. Dose-dependent increases in response have been shown for many nanomaterials~\cite{Iavicoli2018,Ding2005}.

% start a new page without indent 4.6cm
%\clearpage
\end{paracol}
\nointerlineskip
\begin{figure}[H]
\widefigure
\includegraphics[width=0.83\linewidth]{fig1.png}
  \caption{\hl{Gene} %MDPI: please modify hyphen into minus in the image, same as in other figures.
 expression volcano plots for different {cellular models} and TiO\textsubscript{2}-nanobelt concentrations.
   On the x-axis log\textsubscript{2} (fold change) is depicted whereas on the y-axis the \hl{-}%MDPI: please confrim if this should be minus sigh $-$.
log\textsubscript{10} (\emph{p}-value) is depicted. The~dotted lines represent cut-off values for significantly changed genes (log\textsubscript{2} fold change $>$ 0.26 or $<$ $-$0.26, \emph{p}-value $<$ 0.05). A brown color depicts that a gene meets both cut-off criteria, a~blue color relates to meeting only the \emph{p}-value cut-off, an~orange color relates to meeting only the log$_{2}$ fold change cut-off, and a gray color indicates that a gene does not meet any of the criteria.}
\label{fig:fig1}
\end{figure}
\begin{paracol}{2}
%\linenumbers
\switchcolumn
\vspace{-12pt}

\subsection{Pathway Analysis}
Using the differentially expressed genes from the different {cellular models}, an overrepresentation analysis was performed to identify altered pathways after TiO\textsubscript{2}-nanobelt exposure. The~number of significantly overrepresented pathways (\emph{p}-value $<$ 0.05) is shown in Table~\ref{tab:tab1} in the column “Significant”. Concordant to the increase in differentially expressed genes matching our criteria shown in Figure~~\ref{fig:fig1}, the~number of resulting pathways increases with {an increase} in the concentration of TiO\textsubscript{2}-nanobelts. {While} there was a smaller increase in differentially expressed genes, we found a similar increase in resulting pathways for the SAE cell line. There was also an increase in the number of genes found for the {Caco-2/HT29-MTX co-culture; however,} it did not directly translate to an increase in overrepresented pathways. Although~overrepresentation analysis provides a great overview of all processes that are affected, it takes time to go manually over the many overrepresented pathways. Therefore, an~automated method to select desired pathways, i.e.,~toxicity-related pathways, provides a new approach to interpreting the~data.

\begin{specialtable}[H]
\caption{Table listing the number of significantly overrepresented pathways, altered toxicity pathways, and the number of altered toxicity pathways for each~GO-term.}\label{tab:tab1}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcccccc}
\toprule
 & \textbf{Significant} & \textbf{Toxicity} & \textbf{Apoptosis} & \textbf{Inflammation} & \textbf{\begin{tabular}[c]{@{}c@{}}DNA \\ Damage\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Oxidative \\ Stress\end{tabular}} \\ \midrule
\textbf{Pathways} & 1076 & 101 & 66 & 15 & 28 & 2 \\ \midrule
{Caco-2/HT29-MTX} low & 56 & 10 & 5 & 0 & 5 & 0 \\
{Caco-2/HT29-MTX} high & 58 & 10 & 10 & 1 & 0 & 0 \\ \midrule
SAE low & 28 & 3 & 1 & 0 & 2 & 0 \\
SAE high & 50 & 15 & 6 & 1 & 8 & 0 \\ \midrule
THP1 low & 38 & 10 & 9 & 4 & 0 & 0 \\
THP1 high & 201 & 39 & 31 & 9 & 3 & 1 \\\bottomrule
\end{tabular}%
}
\end{specialtable}
\vspace{-12pt}

\subsection{Toxicity-Related Pathways}
To gain more insight into the toxicity-related processes, the~overrepresented pathways were further categorized into pathways related to the apoptotic process, inflammatory response, DNA damage response, and/or oxidative stress. Often this kind of clustering of the pathways is performed manually. However, we implemented a gene-based approach. First, gene sets of four Gene~Ontology~(GO) terms were obtained, i.e.,~“apoptotic process”~(GO\hl{:}%MDPI: please confirm if space should be added after ``:''.
0006915, 1269 genes), “inflammatory response”~(GO\hl{:}0006954, 569 genes),  “cellular response to DNA damage stimulus”~(GO\hl{:}0006974, 762 genes), and~“response to oxidative stress”~(GO\hl{:}0006979, 243 genes). Evidently, these processes are tightly connected. Whereas some can be causative for others, they are expected to overlap. Additional Figure~\ref{fig:figA1} shows the gene overlap between the four GO gene sets in a Venn diagram. Next, we calculated the gene overlap between the pathways from the WikiPathways pathway collection with the annotated genes of the four GO-terms. The~overlap was calculated by dividing the number of toxicity-related genes present in the pathway by the total number of genes present in the pathway. As~a cut-off we used 50$\%$ indicating that at least half of the pathway is directly linked to the GO-term of interest via their gene~overlap.

To select the desired gene overlap cut-off, we compared four cut-offs with each other and an overrepresentation analysis approach to see how many pathways these would give. The~cut-offs we used were 50\%, 60\%, 70\%, and 80\%. For~the overrepresentation analysis we looked for overrepresentation of the pathway genes in the gene lists of the GO-terms. The~results are shown in additional Table~\ref{tab:tabA1}. The~overrepresentation analysis \mbox{(\emph{p}-value~\hl{< $<$}~0.05)} %Please confirm if this is correct.
showed the most pathways to be linked to the four GO-terms; in~this case it showed 401, 250, 203, and 205 pathways overrepresented for the apoptotic process, inflammatory response, DNA damage, and oxidative stress GO-terms, respectively. The~80\% and 70\% gene overlap resulted in almost similar amounts of pathways compared to each other. They even resulted in 0 pathways for the inflammatory response GO-term. Moreover, it stands out that for the gene overlap approach only one pathway meets at least 50\% or higher gene overlap with the oxidative stress GO-term. Additionally, it is interesting to see that an overrepresentation analysis prompts hundreds of pathways compared to just dozens when looking at gene overlap. Comparing the results of the overrepresentation analysis and various cut-offs, the at least 50\% gene overlap cut-off was deemed best for our~approach.

After this filtering step we found that, out of the 1076 pathways in the human pathway collection, 66 pathways are related to the apoptotic process, 15 are linked to inflammatory response, 28 are linked to DNA damage, and 2 are linked to oxidative stress.
It is noticeable that there are only two oxidative stress pathways, i.e.,~Detoxification of Reactive Oxygen Species~(wikipathways\hl{: } %MDPI: added space, please confirm.
WP2824~\cite{WP2824}) and Oxidative Stress~(wikipathways: WP408~\cite{WP408}), that have at least 50$\%$ of their genes overlapping with the genes annotated to the GO-term ``response to oxidative stress''. Based on this finding we can argue that most pathways in the pathway database we have used describe different processes that result in oxidative stress or where oxidative stress is part of/has an influence on the process itself. However, these do not describe oxidative stress as a process.  Nevertheless, these pathways are of interest for our~analysis.

\subsection{Study Effect on Toxicity-Related Pathways}
The number of altered toxicity-related pathways is shown in Table~\ref{tab:tab1}. The~overrepresentation analysis was performed using the differentially expressed genes (log$_{2}$ fold change lower than $-$0.26 or greater than 0.26 and \emph{p}-value lower than 0.05).

The {Caco-2/HT29-MTX co-culture} shows for both concentrations 10 overrepresented pathways. The SAE {cells} show 3 for the low concentration, which could be an indication of a very small toxic response to the TiO\textsubscript{2}-nanobelt exposure, and~15 for the high concentration. The~THP1 cell line shows a clear toxic response activation for the high concentration with 39 overrepresented pathways, and~10 for the low~concentration.
